---
title: "Create POXC plate templates"
author: "Hava Blair"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

## Overview

This R Markdown document takes (1) absorbance data from the plate reader and (2) plate maps typed by hand, and combines them into templates to be used in the next step, `02_poxc_calculation.Rmd` We are creating templates specifically to use with the 'plater' package.

To start:

(1) save absorbance data from plate reader as .xlsx files in raw_data dir. **NOTE** each tab/sheet in your Excel document must be named after the exact plate name as it appears in your `plate_maps.csv` file below. This name is used to join the absorbance data to the correct names.

(2) Type up plate maps as shown in "raw_data/plate_maps.csv".
    All maps in 1 file.
    Make sure plate map names match sheet names in absorbance .xlsx files.
    Save in raw_data dir

There are two ways to use this code: "one-by-one" or "loop".

For "one-by-one" you can call the function make_poxc_plater_template() once for each .xlsx file you wish to turn into plater templates.

For "loop" you make a list of multiple .xslx to be templated and pass it to make_poxc_plater_template() with the walk() function.
See below in "make templates" section.

AFTER running this script: be careful about overwriting templates once you've started to enter "bad wells" info by hand from lab notes.
This would happen in the case where you have already run this script once, but then run it again on the same input file (for whatever reason).
This is easily avoided by renaming the templates folder after you've run the script successfully the first time, or supplying a different dir name in the "make dir if needed" section if you plan to run the script multiple times.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(glue)
library(readr)
library(tidyr)
library(purrr)
library(readxl)
```

## Custom functions

**Thing you might need to change**: depending on when you ran your plates, the specific "rectangle" in the Excel sheet where your absorbance data lives might be different.
This happens in particular when different people adjust the plate reader protocol settings.
Double check your data before running this, and update the value of `excel_plate_rect` below.

Here I'm building up some functions and data structures we'll use to make our templates.
I did it this way because it's pretty common for me to process multiple plates worth of data at the same time, and wrapping this code in function(s) is a way to avoid repeating the lines over and over for each plate.

This code block doesn't return anything explicitly, but it adds the data and functions to your local environment for use later in the process.

```{r}
# does NOT include plate header w/ numbers. DOES include 
# row names with letters. Also includes one empty row (the one
# below H). Should be 8 rows high and 13 columns wide.
excel_plate_rect <- "B24:N32"


# creating an empty template section for marking bad wells 
bad_empty_df <- data.frame(
  "letter" = LETTERS[1:8],
  "1" = rep("", 8),
  "2" = rep("", 8),
  "3" = rep("", 8),
  "4" = rep("", 8),
  "5" = rep("", 8),
  "6" = rep("", 8),
  "7" = rep("", 8),
  "8" = rep("", 8),
  "9" = rep("", 8),
  "10" = rep("", 8),
  "11" = rep("", 8),
  "12" = rep("", 8)
)

colnames(bad_empty_df) <- c("letter",
                            "1",
                            "2",
                            "3",
                            "4",
                            "5",
                            "6",
                            "7",
                            "8",
                            "9",
                            "10",
                            "11",
                            "12")

# gets used in the "make_poxc_plater_temp()" function below
make_template <- function(plate_map, abs_data) {
  # make sure all cols are character
  abs_map <- dplyr::bind_rows(abs_data, plate_map)
  
  abs_map[nrow(abs_map) + 1, ] <- NA
  
  abs_map[nrow(abs_map) + 1, ] <-
    list("bad_wells",
         "1",
         "2",
         "3",
         "4",
         "5",
         "6",
         "7",
         "8",
         "9",
         "10",
         "11",
         "12")
  
  temp <- dplyr::bind_rows(abs_map, bad_empty_df)
  
  temp <- temp %>%
    dplyr::rename(abs = letter)
  
  
  return(temp)
}

# gets used in the "make_poxc_plater_temp()" function below
write_template <- function(plate_name, template) {
  path <- glue::glue("./templates/{plate_name}_plater_template.csv")
  
  readr::write_csv(template, path, na = "")
  
}


make_poxc_plater_temp <- function(target_path) {
  # plate maps -----------------------------------------
  
  maps <-
    readr::read_csv("raw_data/plate_maps.csv", col_types = "cccccccccccccc")
  
  # fill in plate names, drop blank lines, nest by plate
  plate_map_df <- maps %>%
    tidyr::fill(plate_id, .direction = "down") %>% #fill plate name down
    dplyr::filter(!is.na(letter)) %>% # drop blank rows between plates
    dplyr::group_by(plate_id) %>%
    tidyr::nest() %>%
    dplyr::rename(plate_map = data)
  
  
  # absorbance data ----------------------------------
  
  sheets <- readxl::excel_sheets(target_path)
  
  path_sheet_df <- data.frame(path = target_path, sheets = sheets)
  
  # purposely including empty line N32
  abs_df <-
    path_sheet_df %>% dplyr::mutate(abs_data = purrr::map2(
      path,
      sheets,
      ~ readxl::read_excel(
        path = .x,
        sheet = .y,
        # defined at top of this code chunk
        range = excel_plate_rect,
        col_names = c(
          "letter",
          "1",
          "2",
          "3",
          "4",
          "5",
          "6",
          "7",
          "8",
          "9",
          "10",
          "11",
          "12"
        ),
        col_types = "text"
      )
    ))
  
  
  # join and save as csv ------------------------------
  
  abs_map_df <-
    dplyr::left_join(abs_df, plate_map_df, by = c("sheets" = "plate_id"))
  
  
  temp_df <- abs_map_df %>%
    dplyr::mutate(template = purrr::map2(plate_map, abs_data, make_template))
  
  
  plate_names <- temp_df$sheets
  
  template_list <- temp_df$template
  
  purrr::walk2(plate_names, template_list, write_template)
  
  
}

```

## Create folder for templates

Ensures you have a folder named "templates". You don't need to specify a different directory name if you already have a "templates" folder, but you can (if you are testing/changing this script for example, and want to compare outputs).

**Be careful** if you are running this multiple times in one day, that you don't overwrite files after you've started editing them by hand to add "bad well" info from your lab notebook. 


```{r}


# dir = directory = folder
if (dir.exists("templates")){ 
  
  print("templates dir exists, would you like to specify a different dir name?")
  
}else{
  # can change this name if desired
  dir.create(glue("templates"))
}

```


## Make templates from one .xlsx file

```{r}

# provide the path to your excel file w/ absorbance data 
# from the plate reader. 
make_poxc_plater_temp("./raw_data/abs_POXC_CIG_HB_20211217.xlsx")

```

## Make templates from multiple .xslx files

```{r}

# Loop example
# paths to .xlsx files with abs data from plate reader
paths_to_abs_data <- list.files("raw_data",
                                # pulls all files processed
                                # in Dec of 2021
                                pattern = "202112",
                                full.names = TRUE)

# loops through list of paths, saves templates as csv files
walk(paths_to_abs_data, make_poxc_plater_temp) 
```

## Next

-   Go to your lab notebook, review notes, and mark "bad" wells in the plater templates you just created (can mark with any text: X, bad well, x, etc.). Save the updated templates.
-   Be careful if you are testing / running this multiple times with the same input file that you don't overwrite templates you've edited by hand with new blank ones made by this script.
-   Go to `02_poxc_calculation.Rmd`.

## Troubleshooting 

- Made a template, but your plate maps are missing? Double check that the names of the tabs in your absorbance data Excel file **exactly match** the `plate_maps.csv` 
