---
title: "POXC Data Processing"
author: "Hava Blair"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=60))

library(tidyverse)
library(glue)
library(plater)
library(readxl) 

# for my date stamp on exported data
today <- lubridate::today() %>% str_replace_all("-", "")

```

## Overview

Assumes you have already generated `plater` templates with `01_create_poxc_plate_templates.Rmd`

Also assumes that between making the templates and running this script, you have added any relevant info from your lab notebook (bad wells, suspected contamination, etc.) to the "bad_wells" part of the templates you made.

This can be used to run calculations for many plates at once, or you can specify individual plates by the path to their template below in the "Load Data" Section

### Equations

Adapted from Appendix 5 of *Stott, D.E. 2019. Recommended Soil Health Indicators and Associated Laboratory Procedures. Soil Health Technical Note No. 450-03. U.S. Department of Agriculture, Natural Resources Conservation Service.*

To create the standard curve for each plate, determine the slope (b) and y-intercept (a) of a linear regression equation with concentration as the dependent variable (y) and absorbance as the independent variable (x)

conc = a + b\*abs

Active C (mg/kg) = [0.02 mol/L - (a + b \* abs)] \* (9000 mg C /mol) \* (0.02 L solution / mass soil kg)

Where:

-   **0.02 mol/L** is the initial solution concentration\
-   **a** is std curve intercept, **b** is std curve slope, **(a + b \* absorbance)** is the post-reaction concentration\
-   **9000 mg** of C (0.75 mol) is assumed to be oxidized by 1 mol of MnO4 changing from Mn7+ to Mn2+\
-   **0.02 L** is the volume of KMnO4 solution reacted with soil (2 mL 0.2M KMnO4 stock + 18 mL DDI water = 20mL (0.02L) 0.02M solution)\
-   **0.0025 kg** is the weight of soil used (plug in actual mass from weigh sheets)

## Load data

```{r load, message=FALSE, warning=FALSE}

# pull file paths to everything in "templates" folder
# use the "pattern" argument if you want to run specific templates insated of all
tmp_paths <- list.files("templates", full.names = TRUE)

# list of plate names to pass to "read_plates"
# recall that these were extracted from the tab names
# in our original .xlsx data files.
pnames <- basename(tmp_paths) %>% str_replace("_plater_template.csv", "")

# read in the plates, drop bad wells
abs_df <- read_plates(files = tmp_paths, plate_names = pnames, well_ids_column = "wells") %>% 
  mutate(bad_wells = as.character(bad_wells),
    bad_wells = case_when(
    is.na(bad_wells) ~ "ok",
    TRUE ~ bad_wells
  )) %>%
  filter(!is.na(bad_wells)) # drop bad wells


abs_df <- abs_df %>% rename(plate_id = Plate)

sample_mass <- read_csv("raw_data/poxc_soil_mass.csv") %>% 
  mutate(mass_kg = mass/1000) %>% 
  select(-c(weigh_date, person, mass)) %>% 
  rename(id = sample_id)

names <- read_csv("raw_data/cig_names_master.csv") %>% 
  rename(id = sample_id)

```

## Subtract water blank

The Culman KBS protocol indicates to subtract the average of the DDI water blank from all absorbance values before performing any calculations.
The NRCS Tech Note protocol doesn't mention this.
They don't mention it in Pulleman 2021 either, but I suspect that the "Abs_adj" in their equation #1 is adjusted with the DDI blank... otherwise what would it be adjusted for (the paper doesn't specify)?

```{r ddi}
# rounding ddi mean abs to 3 decimal places b/c that's 
# the precision we get from the plate reader
ddi_blanks <- abs_df %>% 
  filter(str_detect(id, "ddi")) %>% 
  group_by(plate_id) %>% 
  summarise(abs_water_blank = round(mean(abs, na.rm = TRUE),3), 
            .groups = "drop") 

# check out range of ddi blanks 
# decide if you need to look more closely at any plates
# update plater templates to remove bad wells
ddi_blanks %>% 
  arrange(abs_water_blank) %>% 
  mutate(plate_id = factor(plate_id), levels = plate_id) %>% 
  ggplot() + 
  geom_point(aes(x = plate_id, y = abs_water_blank)) + 
  coord_flip() +
  ggtitle("DDI Water Blank Absorbance Values") +
  theme_bw()

# what is the mean, sd, cv of the ddi_blanks?
ddi_blanks %>% 
  summarise(mean_abs = mean(abs_water_blank), 
            sd_abs = sd(abs_water_blank), 
            cv_abs = (sd_abs/mean_abs)*100) %>% 
  mutate(across(.cols = everything()), round(., digits = 3))


# join ddi blank values with full dataset
data_long_w_blanks <- left_join(abs_df, ddi_blanks, by = c("plate_id"))

# subtract ddi blank abs from each well, drop ddi blank rows
data_adj_long <- data_long_w_blanks %>% 
  mutate(abs_adj = abs - abs_water_blank) %>% 
  select(-c(abs, abs_water_blank)) %>% 
  filter(!id == "ddi_blank")

```

## Reagent Blanks

The reagent blank value isn't used in any of the calculations, I'm including it here simply as a check on different batches of reagent (note that we made a new batch after the 3/31 run).

```{r reagent}

reagent_blanks <- data_adj_long %>% 
  filter(str_detect(id, "rgt")) %>% 
  group_by(plate_id) %>% 
  summarise(mean_abs_rgt = mean(abs_adj, na.rm = TRUE), 
            .groups = "drop") 

reagent_blanks %>% 
  summarise(mean_abs = mean(reagent_blanks$mean_abs_rgt), 
            sd_abs = sd(mean_abs_rgt),
            cv_abs = (sd_abs/mean_abs)*100) %>% 
  mutate(across(.cols = everything()), round(x = ., digits = 3)) 


  ggplot(data = reagent_blanks) + 
  geom_point(aes(x = plate_id, y = mean_abs_rgt)) +
  coord_flip() +
  ggtitle("Reagent Blank Absorbance Values") +
    theme_bw()



```

## Standard curves

### Select std data and calculate avg abs

```{r std-data}
# std data only, extract KMnO4 conc from sample ids
std_data <- data_adj_long %>% 
  filter(str_detect(id, "M_std")) %>% 
  mutate(id = as.numeric(str_replace(id, "_M_std", ""))) %>% 
  rename(conc = id) 

# calc avg, sd, cv
std_data_avg <- std_data %>% 
  group_by(plate_id, conc) %>% 
  summarise(mean_abs = mean(abs_adj),
            sd_abs = sd(abs_adj), 
            cv_abs = (sd_abs/mean_abs)*100,
            .groups = "drop")

# are we missing any std data? 
std_data_avg %>% 
  filter(is.na(mean_abs)) 

# inspect variability of standard data
# generally HB found <5% CV
std_data_avg %>% 
  ggplot() + 
  geom_histogram(aes(x = cv_abs), bins = 15) +
  ggtitle("CV% of absorbance for standard wells") + 
  labs(subtitle = glue("{nrow(std_data_avg)} observations, n = 3 well replicates each"))

```

### Run linear model and collect model info

Look at the R\^2 values in the table, should be \>=0.99 (or whatever threshold you deem appropriate).

```{r lm}

# create nested structure for modeling
std_nest <- std_data_avg %>% 
  nest_by(plate_id)

# run lm on std data for each plate
# note conc is response var, abs in predictor
# I originally had these switched around
models <- std_nest %>% 
  mutate(model = list(lm(conc ~ mean_abs, data = data)))

# extract model information
model_stats <- models %>% 
  mutate(rsq = summary(model)$r.squared, 
            slope = coef(model)[2], 
            intercept = coef(model)[1]) %>% 
  select(-c(data, model))

model_stats

```

### Inspect standard curve results

```{r curves, echo=FALSE}

max_slope <- round(max(model_stats$slope), 4)
min_slope <- round(min(model_stats$slope), 4)

max_intcpt <- round(max(model_stats$intercept), 4)
min_intcpt <- round(min(model_stats$intercept), 4)

# look at all the std curves together
# identify any problem points for further investigation
ggplot() +
  geom_point(data = std_data_avg,
             aes(x = mean_abs, y = conc, color = plate_id),
             show.legend = FALSE) +
  geom_abline(data = model_stats, aes(
    slope = slope,
    intercept = intercept,
    color = plate_id
  ), 
  show.legend = FALSE) +
  theme_bw() +
  ggtitle("Standard Curves") +
  facet_wrap(vars(plate_id))

# any std curves with R^2 < 0.99?  
bad_curves <- model_stats %>% 
  filter(rsq < 0.99)

# if there are any listed here, need to check those plates
# see if there are obvious bad wells, mark them in template,
# rerun from beginning
bad_curves

```

## Calculate avg abs of samples

Calculate the average absorbance for each sample across the 3 well replicates on the plate.
Join in the linear regression data and soil mass data from above.

```{r avg-abs}
adj_sample_data <- data_adj_long %>% 
  filter(!str_detect(id, "M_std")) %>% # drop std data for now
  group_by(plate_id, id, bad_wells) %>% 
  summarise(mean_abs = mean(abs_adj), 
            sd_abs = sd(abs_adj), 
            cv_abs = ( (sd_abs/mean_abs)*100 ), 
            .groups = "drop")


samples_w_model <- left_join(adj_sample_data,
                             model_stats,
                             by = c("plate_id")) %>%
  select(-rsq) %>% 
  mutate(run_date = as.numeric(str_extract(plate_id, "[:digit:]{8}")))


samples_w_mass <- left_join(samples_w_model, sample_mass, by = c("id", "run_date")) 


# missing any mass data?
samples_w_mass %>% 
  filter(!str_detect(id, "rgt_blank")) %>% # don't need
  filter(is.na(mass_kg)) 

# high CV samples? Look more closely at these
# update bad wells in plater templates for each plate 
samples_w_mass %>% 
  filter(cv_abs >10)

```

## Calculate POXC

Active C (mg/kg) = [0.02 mol/L - (a + b \* abs)] \* (9000 mg C /mol) \* (0.02 L solution / mass soil kg)

-   **0.02 mol/L** is the initial solution concentration\
-   **a** is std curve intercept, **b** is std curve slope, **(a + b \* absorbance)** is the post-reaction concentration\
-   **9000 mg** of C (0.75 mol) is assumed to be oxidized by 1 mol of MnO4 changing from Mn7+ to Mn2+\
-   **0.02 L** is the volume of KMnO4 solution reacted with soil (2 mL 0.2M KMnO4 stock + 18 mL DDI = 20mL (0.02L) solution with concentration 0.02M)\
-   **0.0025 kg** is the weight of soil used (but you should plug in mass from weigh sheets)

```{r calcs}

poxc_calc <- samples_w_mass %>% 
  mutate(poxc_mg_kg = (0.02 - (intercept + (mean_abs * slope))) * 9000 * (0.02/mass_kg))


#add full sample names
named_poxc_calc <- left_join(poxc_calc, names, by = "id") %>%
  filter(id != "rgt_blank") %>% 
  select(-c(
    bad_wells,
    contains("abs"),
    slope,
    intercept
  ))

# save
write_csv(named_poxc_calc, glue("./results/poxc_calculations_{today}.csv"))

```
